---
title: "Quarterly report"
author: "The Real Junk Food Project"
date: "29 June 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = FALSE)
library(ggplot2)
```


This report shows what is possible in terms of automated reported per cafe per quarter (or per week, month or year) with the shift to the continuously collection database.

The first stage is to read-in the data using a secure connection to the database:

```{r}
source("R/sql-read.R")
```

This loads all the datasets we'll be using, including cafes, sources, intercepts and donations. 

Let's see some summary statistics. For all intercepts in this (sample) datasets, the total weight intercepted was:

```{r}
sum(intercepts$weight, na.rm = T)
```

We can put this in plain English for the report, using the following input:

    The total weight intercepted was `r sum(intercepts$Weight, na.rm = T) / 1000000` tonnes.

This results in the following output that the user will see:

The total weight intercepted was `r round(sum(intercepts$weight, na.rm = T) / 1000000, 2)` tonnes.

## By product

We can see the breakdown by food type, for the most frequently intercepted products for example:

```{r}
intercepts_g = group_by(intercepts, Product)
freq_table = intercepts_g %>% summarise(Frequency = n())
top_products = freq_table %>% arrange(desc(Frequency)) %>% top_n(5)
top_products
```

Furthermore, we can plot this output like this:

```{r}
ggplot(top_products) + geom_bar(aes(Product, Frequency), stat = "identity")
```

Or like this:

```{r}
ggplot(top_products) + geom_bar(aes(1, fill = Product, Frequency), stat = "identity")
```

## By source

We can do the same thing for the source of the intercept:

```{r}
intercepts_g = group_by(intercepts, Company)
freq_table = intercepts_g %>% summarise(Frequency = n())
top_companies = freq_table %>% arrange(desc(Frequency)) %>% top_n(5)
top_companies
```

```{r}
ggplot(top_companies) + geom_bar(aes(Company, Frequency), stat = "identity")
```

## By time

By quarter:

```{r}
intercepts_g = group_by(intercepts, Quarter)
freq_table = intercepts_g %>% summarise(Frequency = n())
ggplot(freq_table, aes(Quarter, Frequency)) + geom_bar(stat = "identity")
```

By month:

```{r}
intercepts_g = group_by(intercepts, Month)
freq_table = intercepts_g %>% summarise(Frequency = n())
ggplot(freq_table, aes(Month, Frequency)) + geom_bar(stat = "identity")
```

By week:

```{r}
intercepts_g = group_by(intercepts, Week)
freq_table = intercepts_g %>% summarise(Frequency = n())
ggplot(freq_table, aes(Week, Frequency)) + geom_bar(stat = "identity")
```

By day:

```{r}
intercepts_g = group_by(intercepts, DateIntercepted)
freq_table = intercepts_g %>% summarise(Frequency = n())
ggplot(freq_table, aes(DateIntercepted, Frequency)) + geom_bar(stat = "identity")
```

## Bonus: by food type and week


```{r}
intercepts$Broad_product = "Other"
sel_product = intercepts$Product %in% top_products$Product
intercepts$Broad_product[sel_product] = intercepts$Product[sel_product]
ggplot(intercepts) + geom_bar(aes(Week, fill = Broad_product))
```