---
title: 'The Real Junk Food Project: assessing the environmental and nutritional benefits
  of a global grass-roots movement'
author: "Robin Lovelace and Adam Smith"
date: "April 29, 2015"
output:
  word_document: default
  pdf_document:
    number_sections: yes
    toc: yes
  html_document:
    toc: yes
cls: plos.cls
bibliography: ~/Documents/trjfp.bib
---

```{r, echo=FALSE}
# Todo
# 
```


# Objectives

- To provide an authoritative resource for people wanting to know more about TRJFP,
how it works and what it involves
- To generate 'high level' academic and policy interest in the ideas underlying
the practicalities of TRJFP
- Quantify some of the benefits (e.g. energy) of TRJFP
- To generate some visualisations of data collected by TRJFP
- To explain to people how to set up new TRJFP cafés, the logistics involved and
challenges to overcome
- An anchor point to link to many of the resources that led to TRJFP

# Introduction 

## What is The Real Junk Food Project?

The Real Junk Food Project (TRJFP) is a global movement 
aimed at highlighting and reducing
the costs to people and the environment of
the food industry. Conceptually, TRJFP is a simple and practical campaign to reduce food waste at all levels in the supply chain.
As of .... 2015 TRJFP is also a legal Charitable Foundation
based in Leeds, UK but with operations worldwide (Figure 1).
The highest density of TRJFP cafés and enquiries surrounds its starting point
in West Yorkshire (Figure 2). 

!!! Figure of cafés worldwide

!!! Figure of UK cafés


TRJFP is also 'grass-roots' in that there is no centralised control
structure, in direct contrast to the majority of multi-national food operations.
TRJFP trades from dozens of cafés, restaurants and other outlets
(some of which are mobile) without restrictions on where they. 
There are two defining features of every outlet associated with TRJFP:

1. All food on offer at TRJFP outlets must have been 'intercepted' from the
waste stream. Thus every TRJFP reduces 
the amount of edible food that terminates in landfill.

2. All TRJFP outlets provide food to customers under the principle of
Pay As You Feel (PAYF). No hungry person will be turned away from TRJFP outlets
due to lack of money to pay for food. There is no formal pricing structure.
However, customers are encouraged to pay what they feel the food was worth or
to contribute in other ways, e.g. by assisting in the kitchen.

A slogan of TRJFP that encapsulates these dual criteria is
"feed belies not bins". Thus TRJFP highlights not only the wastage inherent
to modern food systems but also the social impacts of poor food distribution.
TRJFP tackles both sides of the current food crisis: wasteful production and inequitable
distribution and consumption across society.

The relationship between TRJFP 

## Motivations

The growth of TRJFP is organic and ad-hock, driven by motivated people
who decide to set-up their own PAYF outlets where they live. However, TRJFP
business model was carefully designed to maximise its chances of success and
ability to have positive social and environmental benefits in a capitalist
economy.

The twin criteria for TRJFP outlets outlined above were
selected to tackle fundamental flaws of the current model.
On one hand, the reason that people go hungry is not
due to insufficient production of food [@Evans1998].
The problem is access
[@Brinkman2010] and distribution [@Muthayya2013].
On the other, vast quantities of food are wasted [@Stuart2009].

Food waste is ubiquitous at every stage of the supply chain.
The United Nations estimates that
1.3 billion tons is wasted per year, a third of all food [@FAO2013].
The true value could be much higher due to undereporting [@Foresight2011].
This is caused by a wide range of factors: excess use of chemicals
on the fields [@Foley2005]; order cancellations and unharvested
crops due to strict supermarket criteria appearance [@White2011;@Stuart2009];
hoarding and speculation due to
short-term fluctuations in food commodity markets [@Boussard2006];
widespread accidental damage and perishing leading to waste during transit;
inaccurate "sell-by dates" and labelling laws causing premature disposal by supermarkets [@Lipinski2013]
and prolific household food waste [@Household2012].

An excellent overview of the extent of wastage throughout this food supply chain,
"from farm to fork", highlights the problem from a North American perspective
[@Gunders2012]. With globalisation of food systems, these issues
are likely to become increasingly international in scope, as more countries
adopt western-style agriculture and food consumption habits.

The latter source of food waste has received most academic attention,
leading to some insight into its causes. Domestic food waste
is driven by misleading marketing
campaigns such as 'buy on get one free' (BOGOF) offers and inaccurate
expiration dates [@]. The ills of the food system are therefore intimately
entwined with the wider economic system that governs much of
human life, namely global capitalism [@Albritton2009].
Thus *solutions* must tackle not only narrowly-defined aspects of the food
crisis in isolation, but face up to fundamental shortcomings of industrial
food systems. This is the wider motivation for TRJFP.

As outlined above, there has been much
research into the *problems* with the current market-driven food system.
However, there has been a relative dearth of research into potential
*solutions*. This paper is unique in that it is based on real-world experience.

```{r, echo=FALSE}
# rather than speculative 'techno-fixes' [e.g. @ridiculous-solution-a],
# wishful thinking about state-led regulation of f
# or quasi-utopian ideologies. 
# 
# Despite the absolute abundance of food and other raw materials for human
# well-being, a sense of scarcity is produced by the global capitalist system
# [@Eisenstein2011].
```

Critical to all TRJFP outlets is Pay As You Feel (PAYF). No customer is obliged
to pay for food. Everyone has access. However, patrons are encouraged to donate
financially or otherwise (e.g. by volunteering time or skills). This two-way
exchange of value is encourages people to understand the value of the food
services on offer both to themselves and wider society.

TRJFP has been the subject of much commentary.
Despite the large media footprint of the project,
few have taken the time to quantify the 'real' savings that the movement has
produced. By 'real', we refer to changes that can be physically measured. That
is not to discount the importance of the project's ability to change perceptions.
Indeed, encouraging re-evaluation of the food and wider economic systems of
those who interact with TRJFP may be its most important impact. However, these
are impacts that cannot easily be measured.

The focus of this paper is on the subset of changes that TRJFP has brought
about that *can* be measured. Again, this is not to belittle the myriad
intangible changes that have occurred as a result of the ideas inception and
rapid deployment. On the contrary, it is to help put the wider impacts of
TRJFP in perspective. For every measurable impact, for every kilogram of food
not thrown away, it is conceivable that there are a dozen changes in perception
of those directly and indirectly involved in the idea's implementation.
By focussing on what can be measured the paper aims to open the pathway
towards deeper analysis of TRJFP's wider societal impact. In order to focus on
quantifiable impacts, we must begin with the data.

TRJFP is a concept that empowers people from many walks of life to take action
to address entrenched issues with the food system, *in their local areas*.

# The growth of an idea

The first independent PAYF café operating under TRJFP banner opened in Armley, Leeds on December 16th 
2013.^[The first
mention of "The Real Junk Food Project"
in the media was in Good Food magazine, about a PAYF outlet set-up in
Melbourne in early 2013: http://www.goodfood.com.au/good-food/food-news/from-the-rubbish-bin-to-the-restaurant-table-20130527-2n73n.html . It was was the experience of setting-up and running
this operating that led the author (AS) to imagine how the idea could go
global. Inspired by the response from this experience in Australia,
first independent café was set-up shortly after AS returned home to his
home-town in Leeds.]
For the first 4 months this café
opened for 2 hours per day, 2 days per week. Attention from the
international media triggered rapid growth in number of volunteers, 
customers and other visitors. This allowed the café to shift from being a
small operation to opening 7 days per week full-time, within the space of a
couple of months.

The second café was a local community centre.
After a 2 month trial, this project was closed due to internal management issues.
Following continued media interest, the Armley outlet was visited by
individuals from Manchester, Wigan and a number of other cities in the UK
(Fig. x???). There were many enquiries about TRJRP's business model from people
wanting to set-up similar cafés. Enthusiastic responses to the café
from the public, who visited the Armley café in increasing numbers
throughout 2013, also contributed to increased media attention.

Of the cafés to open outside Leeds, the largest in terms of output
has been the Bristol Skipchen, based in Stokes Croft, Bristol.
Co-directors of the Leeds project, Sam Joseph and Catie Jarman,
moved to Bristol to open the venture as part of a campaign against supermarket
waste in 2014. Within days of opening, Bristol Skipchen received
mainstream media attention at the national and international level.
This further boosted interest. From the beginning of 2014 until May 2015,
more than 40 cafés directly or indirectly linked (via correspondence with
AS) opened across the UK.

# Data

```{r, echo=FALSE}
# TODO: write about Birmingham data
```


Data collection was identified as a strategic priority at the
inception of TRJFP. Staff and customers were surprised at the amount
of food that could be taken from the waste stream and converted into
nutritious meals, leading to high demand for further information about
the intercepted food.
Recording of food intercepted became a fundamental practice carried out
by all PAYF cafés in TRJFPs network. At the Armley Junk-tion every gram of food
intercepted is weighed and recorded and other cafés in the network have
(albeit to varying degrees) followed
suit. The information from all outlets is
collated by TRJFP Charitable Foundation, an umbrella organisation,
to be analysed centrally. Along with this, every person fed, every meal made
and all financial donations received during service hours is recorded.
This information is valuable, containing insight into the fundamental changes
required in the food industry. The breakdown of food intercepted by TRJFP
could provide inform strategies to reduce food waste in a more systematic
manner, both in the UK and globally.

The headline figures from this data collection exercise are impressive.
In its first year of service, from December 2014 until December 2014,
the Armley Junk-tion, the first café in TRJFP network,
fed over 10,000 people with more than 23 tons of food intercepted
the waste stream. It is revealing that although the vast majority of
food eaten had officially expired, not a single instance of a negative
reaction to the food provided has been reported.
Thus data from TRJFP suggests that it is safe to eat expired food
and that food wastage could be greatly reduced
by changing laws related to food expiration labels.

## Food interception 

## Food provision to customers

## Financial data



# Results 

## Global distribution and rate of growth

If TRJFP concept is to have global impact, it must be able to expand rapidly
and outstrip the growth-rates of centralised food business models
[e.g. @LondonEconomics2011].^[Although TRJFP
has 'gone viral'
in terms of dissemination ideas and branding this not lead to the
assumption that it is already competitive with the dominant centralised
business model.
It is sobering to remember that TRJFP started from a franchise of
one store in 2014 (notwithstanding other FAYF food initiatives) whereas market
leaders have a head-start of 1000s of outlets and still
continues to grow. McDonald's, for example,
is the largest restaurant franchise in Europe with > 7000 outlets
(> 18,000 worldwide) yet saw rapid growth in sales
(i.e. turnover) and capital expenditure (i.e. new stores) of 17%
between 2007 and 2010 in Europe, against a backdrop of recession.
]
TRJFP-affiliated restaurants have expanded rapidly (Fig. x).
Although growth has been concentrated in Europe and the US, the
brand has received exposure worldwide. The geographical spread of
TRJFP outlets is illustrated in Figure x.

## Waste avoided

Since the inception of TRJFP in 16^th^ December 2013,
statistics collated from all affiliated outlets
show that almost 100 tonnes of food have been intercepted from
the waste stream (67,974 kg to 16^th^ March 2015).

...

## Increased access to nutritious food

Between 16^th^ December 2013 and 16^th^ March 2015, 37,834 separate visits
have been made to TRJFP outlets. 42,180 meals have been served over the same
period. Based on this data combined with the waste avoided

```{r, echo=FALSE}
# 67974 / 42180
```

## Finances and labour

From a business perspective, TRJFP is not a charity in the common sense of the
word. It does not rely on philanthropists making
donations without any expection of personal benefit. Instead, customers are
asked to pay what they feel (PAYF) the food and associated
service is worth. 
This business model has resulted in the donation of £57,097 during the 14
months to 16^th^ March 2015. This equates to an average donation of £1.50
per visit, far below the average price of a meal in the UK of around £5.

```{r, include=FALSE}
57097 / 37834
```

This income is spent on the continuation of TRJFP and is spent on an
outlet-by-outlet bases. Running costs in a typical PAYF café are the same as
running costs in a conventional café, involving st




To illustrate the economics underlying TRJFP even from the perspective
of food wholesalers, consider the case of X, a Fruit and Vegetable wholesale
provider based in West Yorkshire.^[We
do not name the wholesalers due to the commercial sensitivity of the issue.]
X was spending £80,000 per month on disposal of waste food, which consisted
primarily in transport logistics and the UK's Landfill Tax.[See https://www.gov.uk/business-tax/landfill-tax for further details.]
Having heard about TRJFP, X started delivering food deemed
no longer fit for sale but which was still fit for human consumption
to a TRJFP outlet. TRJFP members have also started collecting food from X.
Since this change in business practices, X has reported a 40% saving in
costs associated with food disposal, bringing them down to less than
£50,000 per month.

## Case Study: Armley Junk-tion café

## Energy savings 

# Discussion

- Is it radical to prevent 
- It's not radical it's common sense
- Legal aspects

# Conclusion

Global food systems are highly dynamic, complex and contested. Food is so
embedded in society that no 'golden bullet' solution will be ever be
able to succeed in isolation from wider social change.
The conflicts that arise surrounding food politics have been described
as 'food wars': long-term battles that must be fought on many levels
to overcome vested interests which prioritise short-term profitability
over long-term health and sustainability [@Lang2004]. Specifically,
the solutions "require better processes for making food policies and reform
of the institutions of food governance" [@Lang2004 p. 113].
The solutions are clearly multifaceted [@Godfray2010].

TRJFP and related grass-roots approaches can only provide part of the answer
to the question of how to create a sustainable food system beneficial
to everyone. This paper has demonstrated that TRJFP approach has, however,
had a major positive impact, both in terms of avoidance of food waste and
in terms of providing food to thousands of people, including some of the
most marginalised in society. TRJFP differs radically from the mainstream,
top-down solutions typically proposed in the academic literature by advocating
direct action 'right here right now', rather than campaigning at a distance.
Judging from the public response to TRJFP-affiliated PAYF cafés and the rapid
international growth of outlets, there is appetite for this new approach.

The benefits outlined in this article are large in absolute terms but
pale in comparison to the social and environmental impacts of centralised
food operations. This is not to diminish the TRJFP, the impacts of which
extend far beyond the relatively narrow quantifiable benefits outlined in
this paper. TRJFP has had a vast influence on the way that many people value
food, steering the debate about how we feed ourselves towards contradictions in
the system that are generally hidden from view. Yet the results are decisive:
TRJFP has prevented thousands of tonnes of food being needlessly wasted, in
the process providing nutritional value to the lives of thousands of people.

The fact that this has been possible is itself astonishing and raises the
following questions: How has such continuous wastage of food
become structural to the system? Why have these issues not been
effectively tackled by public policies?
The present analysis of TRJFP cannot answer these
questions, which relate to wider questions about how to ensure equity and
health within a capitalist system [@Jackson2009a]. 
Indeed, the systematic and deliberate destruction of edible food that TRJFP
highlights can be seen simply as one of many indicators of the fundamental
unsustainability of the global economy [@Ehrlich2013; @Martinez-Alier2010].

This paper has demonstrated that decentralised movements such as
TRJFP can empower people to tackle the seemingly intractible issue of food wastage,
on their own terms.
This paper has estimated some of the quantifiable environmental and
nutritional benefits of TJFP, but the wider societal benefits may be far
greater. TRJFP provides a socially empowering alternative to food banks,
a rapidly growing yet flawed solution to food mis-allocation
[@Lambie-Mumford2013].^[Problems with the food bank model include:
the limited supply of tokens regulated by restrictive criteria;
its provision only of non-perishable food, which has low nutritious value;
the fact that some products provided by food-banks require cooking facilities
and skills which some recipients lack; and
the reinforcement of social stigmas related to food inequality by separating
food 'haves' from the 'have nots'.]
Unlike food banks, TRJFP provides a model which encourages involvement by
all members of the community in the solution to food waste. This makes sense
because food mis-allocation is a structural societal problem rather than an
ephemeral issue in need of an short-term technical fix.

# References

